Gradual Soundness: Lessons from Static Python

### 2021-12-21 pj-7.1/paper.tex
pj: checkpoint, 2.2 intro + vtable + chk

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?

@@

Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions
to directly express type checks and type-based optimizations.
These enhancements take full advantage of \SP{} type information;
they are built on top of CPython so that untyped code runs on
the platform.

@@

@@

@@

Cinder adds virtual method tables to typed classes.
``The v-tables are only used for the \code{INVOKE\_METHOD} opcode, which
the static compiler will only emit against a statically known static type.'' -Carl

with vtables, method resolution happens differently than Python: args. get
resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)

vtables are built on Python's vectorcall signature; though new, they're not
totally from scratch

%    No, there’s no overload resolution by arguments in Python. This difference
%    in behavior isn’t so much desired as just a consequence of making
%    `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%    the callable is placed on the stack, then the arguments, then there is a
%    `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%    means that first the callable is resolved, then the arguments. But with
%    `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%    itself, since this gives us opportunity to inline-cache the target of the
%    call instead of always having to call something dynamic and unknown that’s
%    on the stack. So that necessarily implies that first the arguments are
%    resolved and placed on the stack, then the callable is resolved as part of
%    the `INVOKE_METHOD`. -Carl
%
%    2021-11-15: Dino says the JIT will check/upgrade INVOKE_METHOD to INVOKE_FUNCTION
%    2021-07-23: final classes get INVOKE_FUNCTION, not INVOKE_METHOD

@@

As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
come with both a type and an implementation.
The implementation provides the same interface as the built-in;
the main difference is that it adds type checks.

For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
with keys of type \code{K} and values of type \code{V}.
The implementation has three main components:
\begin{itemize}
  \item
    Constructor, basically from dict to CD, checks integrity.
    All CD valuess have to go through this, whether types or untyped.
  \item
    Instances keep a tag, used as exact type in boundary check.
    (Tag may not be CD[i i] but rather a sentinel generated from it.)
  \item
    Writes are protected in every context.
\end{itemize}

%    Each instance of a \code{ChkList[int]} comes with the full type as a tag;
%    run-time checks inspect this tag for an exact match.\footnote{Immutable
%    data structures could use a subtype check. \SP{} does not have any of these
%    because the common Python data structures are all mutable.}

>>>>

%% TODO before submission
%% - look out for PROGRESSIVE TYPES~\cite{pqk-onward-2012}, where changing annotations
%%   helps the static checker find new errors. ChkDict is one example.
%% - idea: formalism would be a SUBSET of Nom, so our burden of proof is low
%% - what are the idiomatic base SP types ... int or Int ?
%% - choose words: static vs typed, nonstatic vs dynamic vs untyped

@@

Cinder is an extension of CPython 3.8 that adds a method-based
JIT compiler, virtual method tables, and several bytecode instructions
to directly express type checks and type-based optimizations.
These enhancements take full advantage of \SP{} type information.
They are built on top of CPython so that untyped code runs on
the platform.

@@

Cinder, the \SP{} runtime system, extends the standard Python runtime in
several ways.
Some of these ensure the integrity of types.
Others use type soundness to generate efficient code.

@@

TODO

@@

Cinder adds virtual method tables (vtables) to typed classes.
These tables help to speed up method dispatch relative to Python's
dynamic lookup.

Specifically, calls to static methods that appear in statically-typed
code use the vtable to find an address for the method.
If the method is part of a final class, then the call is further optimized
to a direct function-call jump.
JIT compilation can also upgrade vtable lookups to function calls.

%% related bytecodes: INVOKE_METHOD INVOKE_FUNCTION
%%  JIT can upgrade I_METH to I_FUN
%%  final classes get I_FUN
%% implementation: built on Python vectorcalls, so it's not totally new
%% beware: with vtables, method resolution is different than Python b/c args. get
%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
%%
%% > This difference in behavior isn’t so much desired as just a consequence of making
%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%% > the callable is placed on the stack, then the arguments, then there is a
%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%% > means that first the callable is resolved, then the arguments. But with
%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%% > itself, since this gives us opportunity to inline-cache the target of the
%% > call instead of always having to call something dynamic and unknown that’s
%% > on the stack. So that necessarily implies that first the arguments are
%% > resolved and placed on the stack, then the callable is resolved as part of
%% > the `INVOKE_METHOD`. -Carl

@@

As mentioned above~(\cref{s:checked-type}), the concrete versions of built-in data structures
come with both a type and an implementation.
The implementation provides the same interface as the built-in,
but uses the type to reject some uses.

For example, the type \code{ChkDict[K, V]} describes a concrete dictionary
with keys of type \code{K} and values of type \code{V}.
The implementation has three main components:
\begin{itemize}
  \item
    A constructor that uses a type and a Python dictionary
    to initialize a checked dictionary.
    The constructor ensures that all values of type \code{ChkDict[K, V]}
    begin with well-typed elements.
  \item
    Guarded update functions.
    Every operation that mutates or extends a checked dictionary must
    validate untyped arguments.
  \item
    A type tag for boundary checks.
    When a checked dictionary enters typed code from an untyped context,
    the runtime checks its key and value types for an exact match.
    For example, a value of type \code{ChkDict[K, V]} matches the type
    \code{ChkDict[str, dynamic]} only if \code{K} is type \code{str}
    and \code{V} is dynamic.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.
